Reconciling agriculture, carbon, and biodiversity in a savanna transformation frontier

### 2016-03-01 paper/main.tex
Supplementary methods draft complete. Required changes to biodiversity-layer.Rmd. Fixed bug in Figure 4 for main text--wrong map selected.
Submitted version 29/2/2016

\title{Reconciling agriculture, carbon, and biodiversity in a savanna transformation frontier}
\author[1,2]{Estes, L.D.}
\author[2]{Searchinger, T.}
\author[2]{Spiegel, M.}
\author[2]{Tian, D.}
\author[3]{Kehoe, L.}
\author[3]{Kuemmerle, T.}
\author[1]{Caylor, K.K.}
\affil[1]{Civil and Environmental Engineering, Princeton University, Princeton, NJ, 08544 USA}
\affil[2]{Woodrow Wilson School, Princeton University, Princeton, NJ, 08544 USA}
\affil[3]{Humboldt University, Berlin}
\date{}

@@

\begin{abstract}
Blah blah
\end{abstract}
%% Text of abstract

@@

%\linenumbers

%% main text
\section*{\large Introduction}

   Meeting growing food demands while minimizing loss of native ecosystems and their carbon and biodiversity presents one of the world’s great challenges for food and the environment.  Many global sustainability studies and climate mitigation pathways call for eliminating or rapidly phasing out emissions from land use change.  CITATIONS IPCC AR V; UK SCIENCE OFFICE.  Unfortunately, these ambitions do not appear realistically attainable in sub-Saharan Africa due to a population likely to more than double to 2.1 billion by 2050. For example, even with healthy linear yield growth from 2006 to 2050 at 3\% to 5\% per year projected by the UN Food and Agriculture Organization (FAO), the region would likely need to add more than 100 million hectares of cropland by 2050.  Searchinger et al. 2013. Much and probably the great bulk of that new cropland is likely to occur in the ~500 million hectares of savannas and shrublands of the region that receive sufficient rainfall to permit crop production.  Searchinger et al. 2013.  This conversion would release tens of millions of tons of carbon and have large impacts on biodiversity.

[I would add in here something about the fact that this is a major challenge, it is one that needs careful planning to achieve, but that this planning has to be done at the country level as this is where most land use policy decisions are made.  Here we look at Zambia, which is perhaps best representative of the challenge.  [this ties us back to the call we made in the NCC paper for finer-level planning, and also inoculates us from the parochiality problem, as we are looking at a representative country].]

Zambia presents an acute part of the challenge.  According to the medium estimate of the United Nations, Zambia’s population is likely to grow 265\% from 16.2 million in 2015 to 43 million in 2050.  (United Nations Department of Economic and Social Affairs Population Division.  2015.  World Population Prospects:  The 2015 Revision, DVD Edition).   As per capita calorie food availability in Zambia is also low, hunger indices high (with child stunting rates at  40\%), https://www.wfp.org/stories/10-facts-about-hunger-zambia GET BETTER CITATION) , and the percentage of the population dependent on agriculture at X\%, Zambia has great need to boost its agricultural production.  Zambia has also been expanding its agricultural exports, and has agronomic and situational characteristics that put it in a strong position to expand exports to other parts of the region.[1]

At the same time, two thirds of Zambia’s 75 milloin hectares is dominated by Miombo Woodland considered dense to be categorized as forest by the UN REDD program, or X\%.  (Vinya, R., Syampungani, S., Kasumu, E.C., Monde, C. \& Kasubika, R. (2011). Preliminary Study on the Drivers of Deforestation and Potential for REDD+ in Zambia. A consultancy report prepared for Forestry Department and FAO under the national UN-REDD+ Programme Ministry of Lands \& Natural Resources. Lusaka, Zambia).  Using this measure of forest, Zambia’s deforestation rate has been estimated at 250,000 to 300,000 hectares per year, among the highest rates of deforestation of any country in the world. Ibid. Fifty percent of these X million hectares have carbon stocks in excess of X tC/h, and an additional X\% have carbon stocks in excess of X.

The biodiversity of these woody savannas is also high.  Overall, the wetter woodland-savannas and shrublands of Africa have a similar average richness for birds and mammals as the world’s wetter tropical forests, and 75\% of these habitats have higher overall vertebrate diversity than 75\% of the rest of the world.   (Searchiner et al. 2015).   Most of Zambia is covered by what the World Wildlife Fund has categorized as Central Zambian Miombo Woodland.  It contains mature miombo woodland trees typically 15 to 20 meters tall, around a quarter of which are evergreen, with a broadleaf shrub and grass understory beneath.[2]  It is also interspersed with a large quantity of wetlands, up to 30\% of many areas.   Its large expanse of habitat make it a center of large mammals, including elephant, black rhino and African buffalo.  Although endemism is low, birdlife is rich with X species, and reptile and amphibian endemism is high, with 19 endemic reptile species 13 endemic amphibians.  LYNDON, CAN WE ADD SOME NUMBERS OF PLANT SPECIES, AND VERTEBRATE SPECIES.   Zambia contains eight other ecoregions according to WWF classifications.[3]   Its extensive wetlands are particular centers of bird and animal life and under extensive pressures.  The value of this habitat is based in part on its remaining relative abundance, which makes it possible for wide-ranging mammals, including many predator species, to survive.

These multiple trends imply challenging trade-offs between agricultural development and environmental concerns.  They raise the question how could decision-makers best target agricultural expansion so that it balances economic, carbon and biodiversity concerns.  A related question is how could a system for targeting well illuminate the trade-offs to support decision-making and a proper role for influence by stakeholders.  And an important question is whether if so what benefits could be achieved by targeting land expansion in certain areas from the standpoint of achieving areas with a desirable mix of objectives.

[We need a somewhat clearer phrasing of our objectives here, or we need to state the objective below the next paragraph. What is that objective: two-fold I would say. 1) Is this sort of modeling approach useful and/or does it have realistic potential for finding compromises in a more general sense (i.e. can it be applied beyond just Zambia), 2) how much potential does Zambia appear to have for reducing the impacts of its expansion. Something along these lines.]

There is a long history of conservation planning that attempts to target conservation areas or objectives while minimizing economic costs subject or while limited land use budget for conservation \cite[e.g.][]{koh_spatially_2010, naidoo_integrating_2006,margules_systematic_2000}. Koh and Ghazoul \citep{koh_spatially_2010} provide one example of an effort to examine trade-offs among production, carbon and biodiversity for oil palm development in Indonesia by giving each land use priority equal weighting.  We adopt a similar approach here, using a newly developed model in which varying weights can be placed on the different objectives, thereby allowing optimal spatial allocations to be found that reflect different degrees of compromise between competing land use interests. To enable this functionality, the model is structured to evaluate each objective in terms of efficiency, or the ratio of cost to yield benefit that results from converting land to agriculture in order to achieve a production target.

We apply this model to answer a basic but important question: can agricultural expansion in Zambia be achieved for substantially lower environmental cost than a business as usual approach, where expansion occurs without consideration of its ecological consequences? This question is particularly important for tropical savanna shrubland, and grassland biomes, which are commonly assumed to provide a low environmental cost reserve for agricultural development \cite{searchinger_high_2015}.

\section*{\large Materials and Methods}
The model we developed, called \emph{AgroEcoTradeoff}, is designed to find the best land use configuration to satisfy the production targets for multiple crops while minimizing four costs: the total land area required, which is inversely related to productivity; the potential losses of carbon and biodiversity; and transportation costs, which are closely correlated with agricultural production costs (ref).  The model takes the basic form:

\begin{equation}
B_{ijk} = \frac{J_{jk}}{P_{ij}} \quad i \in m, \ j = 1, \ldots,n, \ k = 1, \ldots,q 
\end{equation}

Where J$_{k}$ is one of the four costs to be minimized, within a landscape composed of $n$ grid cells containing values for each of the $q$ cost objectives and potential productivity values $P_i$, for each of $m$ crops. $B_{ijk}$ therefore is an efficiency that describes the ecological cost per unit of production, which is normalized so that values are between 0 and 1, so that the different objectives can be compared in common units on a common scale. The normalized values are subtracted from 1 to equate high values with high probability of conversion. Each of these probabilities is then multiplied by a weight (range 0-1, constrained to sum to 1 across all four weights), and the four resulting products are summed, producing a scalarized single objective that represents the landscape conversion probability for each individual crop in each grid cell. An objective function seeks to maximize the sum of these conversion probabilities, based on the requirement that the sum of the potential production of the cells converted for each crop must meet that crop's production target. A second constraint restricts each cell so that it is converted to no more than one crop, as most crops are grown during the same season, and producing multiple crops on the same land would introduce further complexity in setting assumptions about the share of each crop allocated to shared pixels. We therefore formulated a greedy approximation algorithm, based on solving the knapsack problem  \citep{dantzig_discrete-variable_1957} and an algorithm for solving the generalized assignment problem \citep{cohen_efficient_2006}, which allocates each crop to its highest value cells.

This formulation allows quantitatively optimal solutions to be found across a wide range of weight combinations, which is valuable because it is rare to find a single solution that simultaneously obtains the lowest cost for each objective. By finding the optimal solutions across a range of weights, the model can identify the Pareto front, or the set of solutions where the cost in one objective cannot be further minimized without increasing the cost in another. This set then allows model users to choose the most appealing combination of weights based on their land use preferences, and their own assessments of what are and are not acceptable costs within each of the four competing objectives.

This weighting system thus has two basic components.  One reflects the pure value judgment for one output versus another, e.g., how much more a person cares about yield over carbon, or carbon over biodiversity.  The other component expresses each output in quantitative terms so that it reasonably expresses preferences for that output.  For example, units for the carbon, biodiversity and yield outputs are scaled so that someone seeking to weight yield at twice the value each of carbon and biodiversity can do so and have the model find an outcome that reasonably reflects this preference. This approach is necessary to support public decision-making processes, in which different individuals and stakeholders might assign different values to different interests, and where compromise solutions must be sought.

\subsection*{Model inputs} 
For conducting our analysis of potential land use tradeoffs in Zambia, we developed the following spatial datasets at a resolution of 1 km$^2$.

\subsubsection*{\emph{Agricultural potential and production scenarios}}
There are many ways of evaluating agricultural potential factoring in agronomic and economic characteristics.  Here we focused on two crops, maize and soybeans, because of their existing and potential significance for Zambia’s agricultural sector, and because they are considered to be the two dominant commodities that will be grown in Africa's broader savanna regions \citep{searchinger_high_2015}.  Maize is the largest crop by area and consumption in Zambia \citep{fao_fao_2012}, while soybeans are expanding rapidly as a commodity crop to meet rapidly growing regional and global demands for animal feed and direct human consumption \citep{gasparri_emerging_2015}.

We mapped potential yields for each crop using a three-step approach. In the first step, we used the decision support system for agrotechnology transfer \cite[DSSAT, ][]{jones_modelling_2003} to simulate maize and soybean yields at the sites of 40 weather station distributed across Zambia, which represent the points where a 31-year, gridded daily meteorological dataset has received the greatest amount of bias-correction and infilling by weather observations \citep{sheffield_development_2006,estes_changing_2014}. We used soil profiles corresponding to those locations drawn from the WISE v1.1 gridded soil profile database \citep{harvestchoice_converting_2010}, and simulated for each location maize yields over all 31 years under commercial input practices for three-four cultivars representative of short and medium season length according to the Zambian crop calendar \cite{grassini_how_2015}.  This step allowed us to characterize crop responses to Zambia’s inherent climatic variation, and to improved management practices, which was otherwise impossible given the coarse resolution of available observed yield data and the fact that much of Zambia’s current production occurs on low input, smallholder farms (cite). In the second step, we used an empirical model to map the DSSAT-modeled yields onto a 1 km grid, taking advantage of a finer resolution soil map that provides a common frame for estimating both yield potential and soil carbon stocks (see below), but has insufficient detail in several parameters (particularly effective rooting depth) required by DSSAT.  We used a generalized additive model \citep{wood_mgcv:_2001} to predict the values of the DSSAT-simulated yields, using best-fitting subsets of the following weather and soil variables: growing season precipitation; growing degree days; mean growing season shortwave solar radiation; percent organic carbon; pH; percent clay content. The soil predictors represented mean values in the top 1 m of soil, and were extracted from a new 1 km global database of soil properties \cite{hengl_soilgrids1km_2014}.  These first two steps were necessary to 1) capture crop responses to Zambia's climatic variability and potential management practices, and 2) to base potential yields on the same dataset used to estimate carbon costs (see below). This resulted in a map of potential yield under relatively high inputs and good management, and assuming no losses due to other limitations such as diseases. These yield values therefore overestimated what farmers are truly likely to achieve. To correct this, we therefore rescaled the yield maps to match the mean yields (4.5 t/ha for maize and 3.6 t/ha for soybean) that are projected to be achieved in Zambia by 2050, using FAO-projections to derive an annual growth factor, which we then applied to the 2009-2014 reported mean yields for each crop. Overall, the distribution of yields is most important to targeting land use, but the average yield determines the total quantity of land needed to meet production targets.

As a second measure related to agricultural potential, we estimated the travel time to the nearest city of 10,000 people or more (check this), which serves as a proxy for production costs, such as fertilizer prices and transport costs (citation), and is strongly correlated with the suitability of land for farming (cites from ConBio paper).  We used a cost-distance analysis drawing on spatial data for roads, rivers and water bodies, urban centers inside and neighboring Zambia, and national boundaries to calculate travel time in hours. Further details on the cost distance and yield estimation methods can be found in the online electronic supplement.

\subsubsection*{\emph{Carbon}} 
To estimate potential carbon costs from cropland conversions, we developed maps of vegetative and soil carbon stocks for Zambia. Vegetative carbon was calculated from a map of above-ground biomass developed by Baccini et al. \citep{baccini_estimated_2012}, from which we estimated below-ground biomass using savanna and miombo woodland-specific root-shoot ratios \cite{mokany_critical_2006,mugasha_allometric_2013}, and then converted the sum to carbon using a ratio of 0.47 (IPCC 2006 cite). We used the same soil carbon dataset used for crop yield prediction together with an accompanying map of bulk density to calculate soil carbon stocks in the top 1 m. The potential carbon loss due to agricultural conversion was calculated as 100\% of vegetative carbon and 25\% of soil carbon \cite{searchinger_high_2015}.

\subsubsection*{\emph{Biodiversity}}
There are many ways of evaluating biodiversity value and potential impacts, including species richness of different taxa in absolute terms or based on level of threat or endemism, and including factors designed to address the significance of contiguity or the impact of different types of development \citep[e.g.][]{newbold_global_2015,searchinger_high_2015, gasparri_emerging_2015,koh_spatially_2010}. Such measures should reflect both conservation priorities and the quality of available data.  For a fine-scaled, country-level analysis such as this one, the species range map data that form the basis for many impact assessments (Kehoe et al, 2015) have an effective resolution that is too coarse (Laura's cite). We therefore developed a composite measure of biodiversity value based on several datasets. First, we used the potential vegetation maps of East Africa \citep[][]{van_breugel_potential_2011}, a 2010 landcover dataset for Zambia (SERVIR cite), and a shapefile of Zambian protected areas to calculate how much of Zambia's 20 unique vegetation types remains after conversion to croplands and settlements, and what proportion of that remainder falls within national parks. From this we calculated a hybrid measure of rarity and threat, wherein we assigned a weight to each vegetation type based on its relative remaining area in proportion to Zambia's extent, log-transformed to account for a highly skewed distribution, and then re-weighted that weight by each vegetation types protected fraction, following \citep{van_breugel_environmental_2015}. We then used the remaining vegetation map to calculate a second measure, the proportion of untransformed vegetation within a 11X11 km neighborhood centered on each 1 km$^2$, providing a measure of vegetation intactness. By adding this measure to the first index and normalizing, we created a unitless biodiversity score (range 0-1) that gives equal weight to how rare or threatened a habitat is, and how undisturbed it is.

As a final modification, we masked out the areas of existing national parks and game management areas, on the assumption that they would not be used for expanding farmland, but retained a third protected category, forest reserves, as these tend to be far less stringently protected, having had 8.3\% of their area converted to cropland or settlements according to the landcover data. We assigned these areas a high biodiversity score of 0.917 (1 - the already converted proportion), indicating the value of keeping such areas protected, but recognizing that they have some chance of being converted, which can be estimated from their current degree of transformation.

>>>>

\title{Reconciling agriculture, carbon, and biodiversity in a savanna transformation frontier}
\author[1,2]{Estes, L.D.}
\author[2]{Searchinger, T.}
\author[1]{Spiegel, M.}
\author[1]{Tian, D.}
\author[3]{Sichinga, S.}
\author[3]{Mwale, M.}
\author[4]{Kehoe, L.}
\author[4]{Kuemmerle, T.}
\author[1]{Berven, A.}
\author[5]{Chaney, N.}
\author[1]{Sheffield, J.}
\author[1]{Wood, E.F.}
\author[1]{Caylor, K.K.}
\affil[1]{Civil and Environmental Engineering, Princeton University, Princeton, NJ, 08544 USA}
\affil[2]{Woodrow Wilson School, Princeton University, Princeton, NJ, 08544 USA}
\affil[3]{Zambia Agricultural Research Institute, Mt. Makhulu Research Station, Chilanga, Zambia}
\affil[4]{Geography Department, Humboldt University, 10099 Berlin, Germany}
\affil[5]{Program in Atmospheric and Ocean Sciences, Princeton University, Princeton, NJ USA}
\date{}

@@

\begin{abstract}
Rapidly rising populations and likely increases in incomes in sub-Saharan Africa make tens of millions of hectares of cropland expansion nearly inevitable, even with large increases in crop yields. Much of that expansion is likely to occur in higher rainfall savannas, with substantial costs to biodiversity and carbon storage. Zambia presents an acute example of this challenge, with an expected tripling of population by 2050, good potential to expand maize and soybean production, and large areas of relatively undisturbed miombo woodland and associated habitat types of high biodiversity value. Here we present a new model designed to explore the potential for targeting agricultural expansion in ways that achieve quantitatively optimal trade-offs between competing economic and environmental objectives: total converted land area (the reciprocal of potential yield); carbon loss, biodiversity loss, and transportation costs. To allow different interests to find potential compromises, users can apply varying weights to examine the effects of their subjective preferences on the spatial allocation of new croplands and its costs. We find that small compromises from the objective to convert the highest yielding areas permit large savings in transportation costs, and the carbon and biodiversity impacts resulting from savanna conversion. For example, transferring just 30\% of weight from a yield maximizing objective equally between carbon and biodiversity protection objectives would increase total cropland area by just 2.7\%, but result in avoided costs of 27\%-47\% for carbon, biodiversity, and transportation. Compromise solutions tend to focus agricultural expansion along existing transportation corridors and in already disturbed areas. Used appropriately, this type of model could help countries find agricultural expansion alternatives and related infrastructure and land use policies that help achieve production targets while helping to conserve Africa's rapidly transforming savannas.

%Rapidly rising populations and likely increases in incomes in sub-Saharan Africa make tens of millions of hectares of cropland expansion nearly inevitable even with large increases in crop yields.  Much of that expansion is likely to occur in higher rainfall wetter savannas, at substantial costs to biodiversity and carbon storage. Zambia presents an acute example of this challenge with an expected tripling of population by 2050, good agronomic potential to expand maize and soybean production, and large swaths of relatively undisturbed Miombo woodland and associated habitat types of high value to mammals and birds.  Here we present a new model designed to explore the potential to target agricultural expansion in ways that achieve reasonable trade-offs among different economic and environmental objectives.  To allow different interests to find potential compromises, the model allows users to apply different weights to different objectives and examine how they alter outcomes.  We find that modest compromises in yield targets permit large savings in transportation costs, and carbon and biodiversity impacts from savanna conversion. For example, transferring just 30\% of weight from a yield maximizing objective equally to carbon and biodiversity would increase total cropland area by just 2.7\%, but result in avoided cost of 27\%-47\% for the carbon, biodiversity, and transportation cost objectives. Compromise solutions tend to focus agricultural expansion along existing transportation corridors and in already disturbed areas.  Used appropriately, this type of model could offer a practical means by which countries could explore agricultural expansion alternatives and target transportation investments and land use policies to achieve production targets while helping to conserve the values of Africa's savannas.     
\end{abstract}
%% Text of abstract

@@

\linenumbers

%% main text
\section*{\large Introduction}
Meeting growing food demands while minimizing ecological losses presents one of the 21st Century's major challenges.  Many global sustainability studies and climate mitigation pathways call for eliminating or rapidly phasing out emissions from land use change \cite{searchinger_high_2015,field_summary_2014}.  Unfortunately, these ambitions do not appear realistically attainable in sub-Saharan Africa, due to a projected doubling of population by 2050 combined with rapidly growing  economies, which will combine to treble food demand \citep{tilman_global_2011,searchinger_high_2015}. For example, even with the substantial yield growth (3-5\% per year) projected by the UN Food and Agriculture Organization (FAO), the region would likely need to add more than 100 million hectares of cropland by 2050 \citep{searchinger_high_2015}. Much and probably the great bulk of that new cropland is likely to occur in the ~500 million hectares of savannas and shrublands of the region that receive sufficient rainfall to permit crop production \citep{searchinger_high_2015}.

Savannas sometimes receive limited respect in global land use assessments and have been a major focus of agricultural expansion, not only in Africa but in places such as the Cerrado of Brazil and northern Thailand \citep{morris_awakening_2009}.  Their carbon contents and biodiversity values may be completely ignored on the grounds that they are not forests \citep{searchinger_high_2015}. Yet, the wetter woodland-savannas and shrublands of Africa have a similar average richness, although not density, for birds and mammals as the world’s wetter tropical forests, and 75\% of these habitats have higher overall vertebrate diversity than 75\% of the rest of the world \citep{searchinger_high_2015}.  Their carbon contents are also substantial, particularly relative to their likely yields, so that their use for food crops is unlikely to result in fewer carbon emissions per ton of crop \citep{searchinger_high_2015}.  Conversion of over 100 million hectares of these lands would release tens of millions of tons of carbon.

The high likelihood of some cropland expansion in sub-Saharan Africa raises the prospect of whether this expansion could be located in ways that allow agricultural potential to be achieved, but for substantially lower loss of carbon and biodiversity than conventional agricultural development pathways.  Analyzing this question must be done at national or sub-national scales, because that is where key decisions are made, and it requires data that is both in resolution and quality to provide confidence for decision-makers.

In this paper we look at Zambia, which is a bellwether for this land use challenge in sub-Saharan Africa. On the one hand, Zambia has great need to boost its food production. Sixty percent of Zambia's population is rural {\citep{the_world_bank_world_2016}, per capita calorie food availability is just 2100 kcal \citep{fao_faostat:_2016}, and child stunting rates are 40\% \citep{haddad_global_2015}.  According to the medium estimate of the United Nations, Zambia’s population is projected to grow 265\% from 16.2 million in 2015 to 43 million in 2050 \citep{united_nations_world_2015}, which will combine with rapid economic growth \citep[$>$6\%,][]{the_world_bank_world_2016} to substantially increase per capita calorie demands \citep{tilman_global_2011}. Zambia has also been expanding its agricultural exports, and has agronomic and situational characteristics that put it in a strong position to continue to do so to other parts of the region.

At the same time, two thirds of Zambia’s 738,400 km$^2$ land area is dominated by woodland savannas, which have sufficiently dense tree cover to be classified as forest under either of the United Nations Framework Convention on Climate Change thresholds ($>$10 or $>$30\% tree cover) for defining this vegetation type \citep{sexton_conservation_2015}.  These ecosystems store substantial amounts of carbon, with 50\% of Zambia having a carbon density $\geq$130 t ha$^{-1}$ (vegetative carbon and soil carbon in the top 1 m), while an additional 40\% exceeds 80 t ha$^{-1}$. Meanwhile, Zambia’s deforestation rate of 2,500--3,000 km$^2$ year$^{-1}$ is one of the highest in the world \citep{vinya_preliminary_2011}.

The biodiversity of these woody savannas is also high.  Most of Zambia is covered by what the World Wildlife Fund has categorized as Central Zambian Miombo Woodland \citep{olson_terrestrial_2001,hogan_central_2004}. It contains mature miombo woodland trees typically 15 to 20 meters tall, around a quarter of which are evergreen, with a broadleaf shrub and grass understory beneath, and is interspersed with wetlands that comprise up to 30\% of the landscape \citep{hogan_central_2004}. In addition to hosting one of the world's three richest mammal assemblages \cite{olson_terrestrial_2001}, the large expanses of relatively intact habitat make these woodlands, and several of Zambia's eight other ecoregions \citep{fund_ecoregions_2011, estes_zambezian_????}, important centers of large mammal populations, including black rhino, African buffalo, and highly migratory or wide-ranging species, such as elephants and major predators \citep{hogan_central_2004}.  Zambia in general has a high bird diversity \citep{searchinger_high_2015}. Plant diversity in much of the country ranges between 3,000-5,000 species, a richness that is exceeded only in tropical forests \citep{kier_global_2005}.  Although there are relatively few endemic bird and mammal species, reptile and amphibian endemism is high, with 19 endemic reptile species and 13 endemic amphibians \citep{fund_ecoregions_2011}.

There is a long history of conservation planning that attempts to target conservation areas or objectives while minimizing economic costs or within a constrained budget of land area for conservation \cite[e.g.][]{koh_spatially_2010, naidoo_integrating_2006,margules_systematic_2000}. Koh and Ghazoul \citep{koh_spatially_2010} provide one example of an effort to examine trade-offs among production, carbon and biodiversity for oil palm development in Indonesia by giving each land use priority equal weighting.  One of the challenges for land use planning in practice is that decisions involve trade-offs among different objectives and values that most people find difficult to weight against each other \emph{ex ante}, because the values are often abstract, making them harder to evaluate without first viewing the consequential trade-offs.  Here we present a new model in which varying weights can be placed on the competing land use objectives. This set of options then allows model users to choose the most appealing combination of weights based on their land use preferences, and their own assessments of what are and are not acceptable costs within those objectives. The objective of this is to assist different stakeholders in finding compromises between competing land use interests while still meeting agricultural production targets. We apply this model to Zambia so that we can explore its strengths and limitations as a decision-making tool, and also to answer a basic but important question: is it possible to achieve a reasonable balance of different interests by reducing the carbon and biodiversity costs of expansion while still focusing on land use ares with high agricultural potential?

Answering this question is becoming increasingly important as savannas and other grassy biomes receive increasing agricultural development pressure, and may become increasingly targeted as an indirect result of the laudable efforts to conserve tropical forests \citep{searchinger_high_2015}.

\section*{\large Materials and Methods}

The model we developed seeks to find optimal land use configurations that satisfy the production targets for multiple crops while minimizing four costs: the total land area required, transportation costs, carbon released from land conversion, and impacts on biodiversity. Although total land area is treated as a cost, this cost is minimized by selecting areas with the highest likely yields. By minimizing total land required as well as transportation costs, the model finds areas that reflect two important indicators of agricultural potential. By minimizing carbon and biodiversity costs, the model finds areas that minimize two important dimensions of environmental impact.  The model chooses final areas depending on the weights that a user assigns to each of these four objectives.

To enable this functionality, and to identify optimal solutions, the model is structured to evaluate each objective in terms of production efficiency, or the ratio of cost per hectare to crop production per hectare (yield) that results from converting land to agriculture.  For land area costs, output is measured in terms of the number of km$^2$ required to meet the crop production target. For the other objectives, the indicators are transportation costs, carbon loss, or biodiversity impacts per ton of crop yield.

We focus our model initially on two crops, maize and soybeans, because of their existing and potential significance for Zambia's agricultural sector, and their likely dominance of in the future growth of agriculture in Africa's broader savanna regions \citep{searchinger_high_2015}.  Maize is the largest crop by area and consumption in Zambia \citep{fao_faostat:_2016}, while soybeans are expanding rapidly as a commodity crop to meet regional and global demands for animal feed and direct human consumption \citep{gasparri_emerging_2015}. Because the model is designed to meet production targets for multiple crops, we apply algorithms for allocating each crop to its highest value cells.

Further details on model structure and all model input calculations can be found in the online electronic supplement.

\subsection*{Model inputs}

Our model focuses on the costs of cropland expansion into areas that are currently not farmed but could be.  To identify such areas, we calculated the proportion of currently cropped and settled grid cells within each 1 km$^2$ grid cell.  We excluded the proportion of cells that are either protected or with slope greater than 20\% slope, which are both generally considered unsuitable for farming and strong actual predictors of cropland presence or absence \citep{estes_using_2014}.  We estimated the costs of converting each cell as described in the following sections.

\subsubsection*{\emph{Likely yields}}

We spatially estimated likely yields for maize and soybeans using a three-step approach. In the first step, we used the decision support system for agrotechnology transfer \cite[DSSAT, ][]{jones_modelling_2003} to simulate maize and soybean yields at the sites of 40 weather stations distributed across Zambia. They are the points where a 31-year, gridded daily meteorological dataset has received the greatest amount of bias-correction and infilling by weather observations \citep{sheffield_development_2006,chaney_spatial_2014, estes_changing_2014}. We used soil profiles corresponding to those locations drawn from the WISE v1.1 gridded soil profile database \citep{harvestchoice_converting_2010}, and simulated for each location maize yields over all 31 years under commercial input practices for three-four cultivars representative of short and medium season length according to the Zambian crop calendar \citep{grassini_how_2015}.

In the second step, we used an empirical model to map the DSSAT-modeled yields onto a 1 km grid, taking advantage of a finer resolution soil map that provides a common frame for estimating both yield potential and soil carbon stocks (see below), but has insufficient detail in several parameters (particularly effective rooting depth) required by DSSAT to provide a sound empirical basis for running the model directly.  We used a generalized additive model \citep{wood_mgcv:_2001} to predict the values of the DSSAT-simulated yields, using best-fitting subsets of the following weather and soil variables: growing season precipitation; growing degree days; mean growing season shortwave solar radiation; percent organic carbon; pH; percent clay content. The soil predictors represented mean values in the top 1 m of soil, and were extracted from a new 1 km global database of soil properties \cite{hengl_soilgrids1km_2014}.  This two step approach allowed us to more soundly capture crop responses to Zambia's climatic variability and potential management practices, and to base potential yields on the same dataset used to estimate carbon costs (see below).

The first two steps resulted in a map of potential yield under high inputs and good management, and assuming no losses due to other limitations such as diseases. These yield values therefore overestimate what farmers are truly likely to achieve. To reflect more realistic estimates of future yields, we therefore rescaled the yield maps to match the mean yields that are projected to be achieved in Zambia by 2050 (4.4 t ha$^{-1}$ for maize and 3.6 t ha$^{-1}$ for soybean).  We projected these yields by using FAO-projections for sub-Saharan Africa as a whole to derive an annual growth factor and applying that factor to the average 2009-2014 yields for each crop reported in Zambia.  Overall, the distribution of yields derived from our crop modeling is most important to targeting land use, while this choice of average yield determines the total quantity of land needed to meet production targets.

\subsubsection*{\emph{Transportation Costs}}

As a second measure related to agricultural potential, we estimated the travel time to the nearest major town. Travel time serves here as a proxy for differences in output prices and production costs, such as fertilizer prices and transport costs \cite{stifel_isolation_2008}, and, via the density of road networks, is strongly correlated with the suitability of land for farming \citep{laurance_estimating_2015,estes_using_2014}.  We used a cost-distance analysis drawing on spatial data for roads, rivers and water bodies, urban centers inside and neighboring Zambia, and national boundaries to calculate travel time in hours. Total travel costs for any land use scenario equal the total time needed to transport the tonnage of crops produced in each converted cell to the nearest town, which we divide by 20 to produce "truck hours" on the assumption of a 20-ton truck (although the division  of all cell results by this constant does not alter the mapping outcomes).

\subsubsection*{\emph{Carbon}}

To estimate potential carbon costs from cropland conversions, we developed maps of vegetative and soil carbon stocks for Zambia.  Vegetative carbon was calculated from a recent map of above-ground biomass developed by Baccini et al. \citep{baccini_estimated_2012}, from which we estimated below-ground biomass using savanna and miombo woodland-specific root-shoot ratios \cite{mokany_critical_2006, mugasha_allometric_2013}, and then converted the sum to carbon using a ratio of 0.47 \citep{ipcc_land_1996}. We used the same soil carbon dataset used for crop yield prediction together with an accompanying map of bulk density to calculate soil carbon stocks in the top 1 m. The potential carbon loss due to agricultural conversion was calculated as 100\% of vegetative carbon and 25\% of soil carbon \cite{searchinger_high_2015}.

\subsubsection*{\emph{Biodiversity}}

There are many ways of evaluating biodiversity value and potential impacts, including species richness of different taxa in absolute terms or based on level of threat or endemism, and including factors designed to address the significance of contiguity or the impact of different types of development \citep[e.g.][]{newbold_global_2015,searchinger_high_2015, gasparri_emerging_2015,koh_spatially_2010}.  The actual measures selected should reflect both conservation priorities and the quality of available data.  For a fine-scaled, country-level analysis such as this one, the species range map data that form the basis for broader impact assessments \citep[e.g.][]{kehoe_global_2015} have an effective resolution that is too coarse. We therefore developed a composite measure of biodiversity value based on several datasets.

First, we used the potential vegetation maps of East Africa \citep[][]{van_breugel_potential_2011}, a 2010 landcover dataset for Zambia\footnote{downloaded from http://apps.rcmrd.org/landcoverviewer/}, and a shapefile of protected areas to calculate how much of Zambia's 20 unique vegetation types remains after conversion to croplands and settlements, and what proportion of that remainder falls within national parks. From this estimate we calculated a hybrid measure of rarity and threat, wherein we assigned a weight to each vegetation type based on its relative remaining area in proportion to Zambia's extent, log-transformed to account for a highly skewed distribution.  We then re-weighted that weight by the protected fraction of each vegetation type's protected fraction, following \citep{van_breugel_environmental_2015}. We then used the remaining vegetation map to calculate a second measure, the proportion of untransformed vegetation within a 11X11 km neighborhood centered on each 1 km$^2$, providing a measure of vegetation intactness. By adding this measure to the first index and normalizing, we created a biodiversity score (range 0-1 per ha) that gives equal weight to how rare or threatened a habitat is, and how undisturbed its local area.  When the biodiversity score for each converted grid cell converted is summed, it generates a total cost of biodiversity impacts.

As a final modification, we masked out the areas of existing national parks and game management areas to respect the national judgment that they should not be used for expanding farmland.  We retained a third protected category, forest reserves, as these tend to be far less stringently protected, having already had 8.3\% of their area converted to cropland or settlements, but assigned these areas a high biodiversity score.

\subsubsection*{\emph{Potential farmland and production targets}}

Likely future demand for maize and soybeans produced in Zambia is difficult to estimate, and our model can easily be run with different production targets.  For this paper, we adopted targets of a four-fold increase for maize for 2050 compared to 2009-2014 average, and a 10-fold increase for soybeans.  These increases reflect the rate of growth in Zambia from 2000-2014, and a mean production trend for soybean growth in southern Africa estimated by Gasparri et al \citep{gasparri_emerging_2015}, with small additional adjustments to to account for the fact that Zambia has one of the highest agricultural potentials in sub-Saharan Africa and has high potential to expand exports.

\subsubsection*{\emph{Yields on current land}}

The amount of land conversion to meet a production target will depend on the growth of yields on existing croplands, which are well short of their current potentials \citep{waddington_getting_2010}.  For the principal scenarios we present in this paper, we assume that yields on current cropland will achieve the average likely yields for Zambia's \emph{potential} farmland.  However, to examine the possible significance of yield gains on existing cropland, we also examined scenarios in which the yield gains on existing cropland 1) only manage to close 50\% of the gap between current and our projected future yields, and 2) close 150\% of the gap.

\subsection*{Normalization and Weighting}

Weighting different objectives is to some extent based purely on preferences, but it is important that these be expressed in a numerical form and processed so that a quantitatively optimal solution for that set of preferences is identified. To achieve this result, we first express the different objectives as efficiencies (e.g. carbon loss per ton of crop yield, as described above), and then normalize the units of each objective based on the range of its scores. We then multiplying each normalized objective by its corresponding weight (which fall between 0 and 1 and must sum to 1, with values closer to 1 indicating stronger preferences) and then adding the resulting products. The result is a single score that ranks each grid cell according to its ability to meet the production target for the lowest cost, given those weights. The model then selects cells in descending rank order until each the cumulative production of converted cells reaches each crop's corresponding target (see online electronic supplement for details of land allocation between crops).
